Posterior summarisation in Bayesian phylogenetics using Tracer 1.7

### 2018-04-12 manuscript/response.md

We again thank the Reviewer for this bug alert and will attempt to fix in a future release.
**Test loading a too large file.**

# Associate Editor's comments

*Almost all of the material is clear and understandable, even to readers who have never used the software. For me the exception was the paragraph starting at 66. Since this appears to be a new feature, it probably could be elaborated on. Some general description of the types of things that can be conditioned on (models, the presence of variables in models, ...) and how the software does this, possibly with a figure, might help before or with some of the specifics on lines 68-79; I didn't actually see much about this in briefly looking at the tutorials on beast.community/tracer.*

We agree that our original description of visualising conditional posteriors was confusing, chiefly because we failed to provide instructions on how a user can generate the plots.  We have rectified this short-coming by adding these instructions.  The beginning of the paragraph now reads:

"Tracer offers a solution of visualising conditional posterior distributions as well.
Selecting one continuous and one categorical parameter generates side-by-side violin or boxplots under the Joint-Marginal panel.  These plots present the continuous parameter distribution conditioned on the observed categorical values.  A typical use case involves Bayesian stochastic search variable selection..."

The remainder of the paragraph provides several use-cases, in line with the Associate Editor's suggestion.

### 2018-04-12 manuscript/response.md

We again thank the Reviewer for this bug alert and will attempt to fix in a future release.
**Test loading a too large file.**

# Associate Editor's comments

*Almost all of the material is clear and understandable, even to readers who have never used the software. For me the exception was the paragraph starting at 66. Since this appears to be a new feature, it probably could be elaborated on. Some general description of the types of things that can be conditioned on (models, the presence of variables in models, ...) and how the software does this, possibly with a figure, might help before or with some of the specifics on lines 68-79; I didn't actually see much about this in briefly looking at the tutorials on beast.community/tracer.*

Please see our response to Reviewer #1 (S1).  The remainder of the quoted paragraph provides several use-cases, in line with the Associate Editor's suggestion.

### 2018-04-14 manuscript/response.md

We thank the Reviewer for this encouragement.

## Regarding the software:

*S1: How does the conditional posterior visualization work? That seems like a major improvement, but it’s not shown anywhere or described beyond a mention in the intro. The documentation basically says “try things out!”-- this mostly works except for here.*

We agree that our original description of visualising conditional posteriors was confusing, chiefly because we failed to provide instructions on how a user can generate the plots.  We have rectified this short-coming by adding these instructions.  The paragraph now reads:

"Tracer offers a solution of visualising conditional posterior distributions as well.
Selecting one continuous and one categorical parameter generates side-by-side violin or boxplots under the Joint-Marginal panel. These plots present the continuous parameter distribution conditioned on the observed categorical values. A typical use case involves Bayesian stochastic search variable selection (BSSVS), a form of model averaging, in which parameters influence the likelihood function only when a specific model is selected by a random indicator function."

@@

At the request of the Reviewer, we have restored this feature in the software. See [Issue #146](https://github.com/beast-dev/tracer/issues/146).

*S5: Please update the link on http://beast.community/tracer , which currently links to 1.6 (even though it says it’s linking to 1.7)*

Done.

*P1: People mostly use Tracer to diagnose convergence problems. I think the best use of a paper about Tracer would be to walk readers through a side-by-side comparison of two runs, one of which has converged and another that hasn’t. This is additional work, but I think it would be very valuable to guide the community in making such calls. At least, there should be a link in the paper to some material giving such an example. The current Figures 1 b-e and 2a-c feel a little generic-- we know what these plot types look like.*

@@

*Title: We cannot resist the temptation to suggest to the more active “Summarizing Bayesian phylogenetic posterior distributions using Tracer 1.7”*

We elect to keep our original title "Posterior summarisation in Bayesian phylogenetic using Tracer 1.7" as it is shorter and reads more naturally in the form (adjective-noun) -> conjunction -> (adjective-noun) -> conjunction -> (noun-modifier).

*L68: the BSSVS description doesn’t really feel like it comes to a satisfactory conclusion. Can you more clearly describe how Tracer helps with this type of analysis? Is the sentence at line 80 a continuation of this thought, or starting something new?*

As stated in a previous response, we have rewritten this paragraph to better describe the option in Tracer to visualise conditional posteriors.
The sentence at line 80, starting a section on demographic plot generation/visualisation, has been moved to a new paragraph.

@@

*1) It would be nice to export a csv file with the parameter estimates (mean, 95% HPD,  etc) and ESS values so that this can be used easier in reports.*

We thank the Reviewer for this suggestion.  It is implemented under the File menu -> Export Data Table.

*2) There seems to be some memory issue in Tracer when multiple large files are opened. Closing and opening gets around the issue. It would be nice if that is fixed in some future release.*

We again thank the Reviewer for this bug alert, which is related to the default memory allocation in Tracer. We have increased this default substantially, to decrease the odds the Tracer users will encounter this problem. Should users still encounter this problem, they can increase memory allocation as well, which we now discuss in one of the Tracer tutorials.

@@

*-Fig 1(b)-(d): If you want to make space for additions, two of these could be cut and simply described as a list of (very familiar) available graphics for comparing multiple (marginal) distributions. By contrast, Fig 2(b)-(d) are less familiar and really need a figure to accompany the text description.*

Please see our response to Reviewer #1 (P1).

*-l67: Something like the sentence starting on l71 could be moved here to indicate generally how the conditional distributions are constructed from the sampled parameter values. It would be valuable to say something about how generally the indicator functions can be defined, how they are selected...*

As stated in a previous response, we now provide proper instructions on how to visualise conditional posteriors.

### 2018-04-14 manuscript/tracer17_systbiol.tex
briefly handle condition distributions
conditional plots can use integers as well.

\address{
$^{1}$Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK;
$^{2}$Department of Computer Science, University of Auckland, Auckland, NZ;
$^{3}$Centre for Computational Evolution, University of Auckland, Auckland, NZ;
$^{4}$Department of Microbiology and Immunology, Rega Institute, KU Leuven, Leuven, Belgium;
$^{5}$Department of Human Genetics, University of California, Los Angeles, USA;
$^{6}$Department of Biostatistics, University of California, Los Angeles, USA \\[6pt]
$^{\ast}$Correspondence to:
\href{a.rambaut@ed.ac.uk}{\url{a.rambaut@ed.ac.uk}},
\href{alexei@cs.auckland.ac.nz}{\url{alexei@cs.auckland.ac.nz}}
and
\href{msuchard@ucla.edu}{\url{msuchard@ucla.edu}}}

@@

\tracer\ examines the posterior samples from all the available parameters, treating continuous, integer and categorical parameters appropriately,
from a trace
and presents statistical summaries and visualizations.
Further, \tracer\ can analyse a single trace or combine samples from multiple files.
Immediately apparent in the default \tracer\ view, the effective sample size (ESS) is one such statistic that
allows users to assess the number of effectively independent draws from the posterior distribution the trace represents (Figure \ref{fig:overview}a).
Colour coding assists the user in determining potential MCMC mixing problems, with arbitrary cut-off values at 100 and 200.

@@

\begin{itemize}[leftmargin=*]
\vspace{-0.5em}
\item Estimates: Reports common summary statistics such as the sample mean, standard deviation, highest posterior density interval and ESS.  Also presents a histogram of sample values for a single selected parameter (Figure \ref{fig:overview}a) or  side-by-side boxplots for multiple continuous parameters (Figure \ref{fig:overview}b).
\item Marginal Density: Draws density plots for the selected parameter(s), including  kernel density estimates (Figure \ref{fig:overview}c), histograms and violin plots (Figure \ref{fig:overview}d) for continuous parameters and frequency plots for categorical or integer parameters.
\vspace*{-0.25em}
\item Joint-Marginal: Visualisation in this panel appears after selecting two or more parameters and the plot form depends on the parameter types. We show several examples in the next section of the paper.
%
\begin{figure}[t]
\subfigure[Tracer overview]{\includegraphics[width=0.485\textwidth]{figure_1a_rabv-overview.pdf}} \\
\subfigure[Boxplot]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1b_rabv-boxplots.pdf}}
\subfigure[Density]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1c_rabv-density.pdf}}\\
\subfigure[Violin]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1d_rabv-violin.pdf}}
\subfigure[Trace]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1e_rabv-traces.pdf}}
\vspace{-0.25cm}
\caption{Overview of \tracer\ functionality and individual parameter visualisations: (a) Main \tracer\ panel upon loading a single trace file; (b) boxplot representation of two continuous parameters; (c) kernel density estimates; (d) violin plots; (e) the actual traces connecting the parameter values visited by the Markov chain.}
\label{fig:overview}
\end{figure}
%

@@

\tracer\ offers a solution of visualising conditional posterior distributions as well.
A typical use case involves Bayesian stochastic search variable selection (BSSVS), a form of model averaging, in which parameters influence the likelihood function only when a specific model is selected by a random indicator function.
Under BSSVS, a posterior estimate of the parameter should only sample values from states where the indicator equals one.
Discrete phylogeographic analyses frequently employ BSSVS due to the potentially large amount of transition rates that need to be estimated \citep{Lemey2009}, but this is also relevant when employing model averaging approaches, e.g., over relaxed molecular clocks \citep{Li2012}.

>>>>

\address{
$^{1}$Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK;
$^{2}$Department of Computer Science, University of Auckland, Auckland, NZ;
$^{3}$Centre for Computational Evolution, University of Auckland, Auckland, NZ;
$^{4}$Department of Microbiology and Immunology, Rega Institute, KU Leuven - University of Leuven, Leuven, Belgium;
$^{5}$Department of Human Genetics, University of California, Los Angeles, USA;
$^{6}$Department of Biostatistics, University of California, Los Angeles, USA \\[6pt]
$^{\ast}$Correspondence to:
\href{a.rambaut@ed.ac.uk}{\url{a.rambaut@ed.ac.uk}},
\href{alexei@cs.auckland.ac.nz}{\url{alexei@cs.auckland.ac.nz}}
and
\href{msuchard@ucla.edu}{\url{msuchard@ucla.edu}}}

@@

\tracer\ examines the posterior samples from all the available parameters - treating continuous, integer and categorical parameters appropriately - from a trace and presents statistical summaries and visualisations.
Further, \tracer\ can analyse a single trace or combine samples from multiple files.
Immediately apparent in the default \tracer\ view, the effective sample size (ESS) is one such statistic that
allows users to assess the number of effectively independent draws from the posterior distribution the trace represents (Figure \ref{fig:overview}a).
Colour coding assists the user in determining potential MCMC mixing problems, with arbitrary cut-off values at 100 and 200.

@@

\begin{itemize}[leftmargin=*]
\vspace{-0.5em}
\item Estimates: Reports common summary statistics such as the sample mean, standard deviation, highest posterior density interval and ESS.  Also presents a histogram of sample values for a single selected parameter (Figure \ref{fig:overview}a) or  side-by-side boxplots for multiple continuous parameters (Figure \ref{fig:overview}b).
\item Marginal Density: Draws density plots for the selected parameter(s), including  kernel density estimates (Figure \ref{fig:overview}c), histograms and violin plots (Figure \ref{fig:overview}d) for continuous parameters and frequency plots for categorical or integer parameters.
\vspace*{-0.25em}
\item Joint-Marginal: Visualisation in this panel appears after selecting two or more parameters and the plot form depends on the parameter types. We show several examples in the next section of the paper.
%
\begin{figure}[t]
\subfigure[Tracer overview]{\includegraphics[width=0.485\textwidth]{figure_1a_rabv-overview.pdf}} \\
\subfigure[Boxplot]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1b_rabv-boxplots.pdf}}
\subfigure[Density]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1c_rabv-density.pdf}}\\
\subfigure[Violin]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1d_rabv-violin.pdf}}
\subfigure[Trace]{\includegraphics[width = .23\textwidth, height = 3cm]{figure_1e_rabv-traces.pdf}}
\vspace{-0.25cm}
\caption{Overview of \tracer\ functionality and individual parameter visualisations: (a) Main \tracer\ panel upon loading a single trace file; (b) boxplot representation of two continuous parameters; corresponding (c) kernel density estimates; (d) violin plots; and (e) the actual traces connecting the parameter values visited by the Markov chain.}
\label{fig:overview}
\end{figure}
%

@@

\tracer\ offers a solution of visualising conditional posterior distributions as well.
Selecting one continuous and one integer or categorical parameter generates side-by-side violin or boxplots under the Joint-Marginal panel.
These plots present the continuous parameter distribution conditioned on the unique integer or categorical values.
A typical use case involves Bayesian stochastic search variable selection (BSSVS), a form of model averaging, in which parameters influence the likelihood function only when a specific model is selected by a random indicator function.
Under BSSVS, a posterior estimate of the parameter should only sample values from states where the indicator equals one.
Discrete phylogeographic analyses frequently employ BSSVS due to the potentially large amount of transition rates that need to be estimated \citep{Lemey2009}, but this is also relevant when employing model averaging approaches, e.g., over relaxed molecular clocks \citep{Li2012}.
